\begin{thebibliography}{99}

\bibitem{A} L. Ammerall, {\bfseries Computer Graphics for Java
Programmers}, John Wiley \& Sons, 1998.

\bibitem{AMR} U. M. Ascher, R. M. M. Mattheij, and R. D. Russell,
{\bfseries Numerical Solution of Boundary Value Problems for Ordinary
Differential Equations}, SIAM, 1995.

\bibitem{AG} D. Assaf, IV, and S. Gadbois, Definitions of Chaos,
{\bfseries The American Mathematical Monthly}, {\bfseries 99}, No.\ 9, p. 865.

\bibitem{BBCDS} J. Banks, J. Brooks, G. Cairns, G. Davis and
P. Stacey, On Deveney's Definition of Chaos, {\bfseries The American
Mathematical Monthly}, {\bfseries 99}, No.\ 4, pp.\ 332-334.

\bibitem{BA} R. G. Bartle, {\bfseries The Elements of Real Analysis, 2$^{nd}$
Edition}, John Wiley \& Sons, 1976.

\bibitem{BT} J. L. Buchanan and P. R. Turner, {\bfseries Numerical Methods
and Analysis}, McGraw-Hill, Inc., 1992.

\bibitem{BF} R. L. Burden and J. D. Faires, {\bfseries Numerical
Analysis, $5^{th}$ ed.}, PWS-KENT Publ.\ Comp., 1993.

\bibitem{B} J. C. Butcher, {\bfseries The Numerical Analysis of
Ordinary Differential Equations: Runge-Kutta and General Linear
Methods}, Wiley, 1987.

\bibitem{CSD} S. D. Conte, The numerical solution of linear boundary
value problems, {\em SIAM Review}, {\bfseries 8}, 1966, pp.\ 309--321.

\bibitem{CdB} S. D. Conte and C. de Boor, {\bfseries Elementary
Numerical Analysis}, McGraw-Hill, 1980.

\bibitem{CT} J. W. Cooley and J. W. Tukey, An algorithm for the
machine calculation of complex Fourier series, {\bfseries Math. Comp.},
vol.\ {\bfseries 19}, 1965, pp.\ 297--301.

\bibitem{D} R. L. Devaney, {\bfseries An Introduction to Chaotic
Dynamical Systems}, The Benjamin/Cummings Publ.\ Co.\ Inc., 1986.

\bibitem{GvL} G. H. Golub and C. F. van Loan, {\bfseries Matrix
Computation, $3^{rd}$ ed.}, John Hopkins Univ.\ Press, 1996

\bibitem{HNW} E. Hairer, S. P. Norsett and G. Wanner, {\bfseries
Solving Ordinary Differential Equations I: Nonstiff Problems},
Springer-Verlag, 1991.

\bibitem{HW} E. Hairer and G. Wanner, {\bfseries Solving Ordinary
Differential Equations II: Stiff and Differential-Algebraic Problems},
Springer-Verlag, 1996.

\bibitem{He} P. Henrici, Fast Fourier methods in computational complex
analysis, {\em SIAM Review}, {\bfseries 21}, 1979, pp.\ 481--527.

\bibitem{HS} Morris W. Hirsch and Stephen Smale, {\bfseries Differential
Equations, Dynamical Systems, and Linear Algebra}, Academic Press,
1974.

\bibitem{WH} Wolfgang Hackbusch, {\bfseries The Concept of Stability
in Numerical Mathematics}, Springer-Verlag, 2014.

\bibitem{I} A. Iserles, {\bfseries A First Course in the Numerical
Analysis of Differential Equations}, Cambridge Univ.\ Press, 1996.

\bibitem{IK} E. Issacson and H. B. Keller, {\bfseries Analysis of
Numerical Methods}, Dover Publ.\ Inc., 1994.

\bibitem{KC} D. Kincaid and W. Cheney, {\bfseries Numerical Analysis,
Mathematics of Scientific Computing, $3^{th}$ ed.}, Brookes/Cole,
2002.

\bibitem{K} H. B. Keller, {\bfseries Numerical Methods for Two-Point
Boundary-Value Problems}, Dover Publ.\ Inc., 1992
  
\bibitem{La} J. D. Lambert, {\bfseries Computational Methods in
Ordinary Differential Equations}, Wiley, 1973.

\bibitem{Lb} J. D. Lambert, {\bfseries Numerical Methods for Ordinary
Differential Systems}, Wiley, 1991.

\bibitem{O} J. M. Ortega, (1968). "The Newton-Kantorovich Theorem",
{\bfseries The American Mathematical Monthly}, {\bfseries 99}, No.\ 4,
pp.\ 658–660.

\bibitem{PJS} H.-O. Peitgen, J. Hartmut and S. Dietmar, {\bfseries
Chaos and Fractals: New Frontiers of Science}, Springer-Verlag, 1992.

\bibitem{R} Walter Rudin, {\bfseries Principle of Mathematical Analysis,
3$^{rd}$ Edition}, McGraw-Hill, 1978.

\bibitem{S} L. F. Shampine, {\bfseries Numerical Solution of Ordinary
Differential Equations}, Chapman \& Hill, 1994.

\bibitem{GS} G. F. Simmons, {\bfseries Differential Equations with
Applications and Historical Notes}, McGraw-Hill, 1972.

\bibitem{St} G. Strang, {\bfseries Introduction to Applied
Mathematics}, Wellesley-Cambridge Press, 1986.

\bibitem{Smi} G. D. Smith, {\bfseries Numerical Solution of Partial
Differential Equations, 2$^{nd}$ Edition}, Oxford University Press, 1978.

\bibitem{Str} W. A. Strauss, {\bfseries Partial Differential
Equations: an Introduction}, John-Wiley \& Sons, 1992.

\bibitem{SH} A. M. Stuart and A. R. Humphries, {\bfseries Dynamical
Systems and Numerical Analysis}, Cambridge Univ.\ Press, 1996.

\end{thebibliography}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "notes"
%%% End: 
